---
title: "R Notebook"
output: pdf_document
---

```{r}
getwd()
countries<-read.table("./ecosystems.txt",sep="\t",header=T)
# countries<-read.table("./2nda_parte/FANGA_TAUFA_del_tab.txt",sep="\t",header=T)
str(countries)

countries2 <- scale(countries)

## Matriz de interdistancias (Parte 1)
countries.D1 = dist(countries2, method = "euclid")

#clustering
require(cluster)
clusterW<-hclust(countries.D1^2, method="ward.D2")

#Agglomerateive clustering, UPGMA
plot(clusterW, hang=-1, labels = countries[,1])
```

```{r}
require(cclust)
results<-data.frame()
for(x in c (1,2,3,4,5,6,7,8,9,10)){
  result.km.x = kmeans(countries2, centers=x, nstart=1000)
  Silh.km.x<-silhouette(result.km.x$cluster,dist(countries2))
  results[(x),1]<-x
  results[(x),2]<-result.km.x$totss
  results[(x),3]<-result.km.x$tot.withinss

  if (x < 6) {
    PseudoF.km.x<-clustIndex(result.km.x, countries2, index="calinski")
    results[(x),4]<-PseudoF.km.x
  } else {
    results[x,4] <- NA
  }
  
  if (x == 1) {
    results[(x),5]<- NA
    results[(x),6]<- NA
  } else {
    Overall.Silh.km.x<-mean(Silh.km.x[,3])
    results[(x),5]<-Overall.Silh.km.x
    results[(x),6]<- (results[(x-1),3] - results[(x),3]) / results[(x-1),3]
  }
  if (x != 1) plot(Silh.km.x)
}
colnames(results)<-c("K","TOTSS","WITHINSS","PseudoF","AvgSilh","DeltaTESS")
results
```

One can see clearly (mainly from the silhouettes) that a small number of clusters is not going to give a great classification of the data, but a big number of clusters won't be really robust given the small amount of data that we have. If we need to choose between 2 or 3, probably choose 2, silhouettes seem better (when looked at it on every point) also has better delta tess and slightly better pseudoF. PseudoF crashes for over 6 groups cuz clustIndex programmers are not able to handle single point groups :') block below can do it properly

```{r}
# Compute K-means for a range of values of K
require(vegan)
result.cascadeKM = cascadeKM(countries2, inf.gr=2, sup.gr=10, iter = 1000, criterion ="calinski")
attributes(result.cascadeKM)
result.cascadeKM$results
result.cascadeKM$size
```

```{r}
result.km.2 <- kmeans(countries2, centers=2, nstart=1000)
clusplot(countries2, result.km.2$cluster, main = "Kmeans plot, k = 2", color = TRUE, labels=2)
```

```{r}
result.km.3 <- kmeans(countries2, centers=3, nstart=1000)
clusplot(countries2, result.km.3$cluster, main = "Kmeans plot, k = 3", color = TRUE, labels=2)
```

As before, all seems to indicate 2 clusters is better than 3. The plot is way nicer, clusters are separated, etc.