---
title: "Practical 4"
author: "Eudald Romo Grau y Laura Santulario Verdú"
date: "March 15, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


__1.__ Load the contingency table in the R environment.  

```{r}
mum_child <- read.table("Mother_child.txt", header = TRUE, sep = " ", dec = ".", row.names = 1)
colnames(mum_child) <- c("<18", "18-25", "25-30)", "[30-35)", "35")
```


__2.__ (2p) Is there association between birth weight and age of the mother? Perform a chisquare test for independence between row and columns. Report the chi-square statistic, degrees of freedom, and p-value.  

##Chisquare test  

```{r}
chisq.test(mum_child)
```

Aplaying the independece test to our data: $$H_0: \text{The mother's age and the weight of the child at birth is independent}$$ 
$$H_1: \text{The mother's age and the weight of the child at birth is dependent}$$

We have obtained a *p-value = 8.046e-10*, what means that the probability of rejecting $H_0$ being true is 8.046e-10 (nearly 0). So considering an $\alpha$ level equal to 0.05 (0.05 >> 8.046e-10) we can reject with enough statistical significance the null hypothesis. So there is dependece between the age of the mum and the weight of the babies.  


__3.__ (2p) Compute the correspondence matrix. Which category is the most frequent category? Which the less frequent?  

```{r}
P <- round((1/sum(mum_child))*mum_child, 3)
P
which(P == max(P))
which(P == min(P))
```

The most frequent category is that which the mothers are between 18 and 25 years and the babies have born with a weight between 2 and 3 kg.  

On the other hand the less frequent categories are that where the mum is between 18 and 25 years and the baby weight less than 1 kg (what means that is not often that a mum of this age has premature babies) 
The mums older than 35 years with babies with 1-2 kg.  

__4.__ Compute the expected counts under the assumption of independence. What is the sum of all expected counts?








